\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-20 coqimpl.tex
Finishing Language Design Section. Try to fix graph embedding and format.
Comments Reply.

\textbf{Family compiled into module.} This is the main idea of compiling family components into Coq's primitives: we compile family into coq module, family type into module type and context of a given term into parameters of the module. Unlike module, we need to faithfully reflect the \textit{late binding} nature of family : for each defining field $ .. \cL\sigma \vdash t : T $, we need to compile $t$ using ``universal-quantifier-wrapped term'' : $.. \vdash \lambda t : \forall \cL \sigma. T$. However, instead of using Coq's universal quantifier, we choose module and module parameters to achieve this wrapping---each family type in the context of the judgement will be one module parameter with the compiled module type, and thus each field of the family will compile into a functor parametrized by its context. Doing so we can also get rapid feedback from type checking when the users are defining each field interactively with Coq.
% Insert one pseudo-code example for explanation of the mechanism
\begin{figure}
  \begin{minipage}[t]{0.25\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Family A.
Field a : nat 
  := 1.
Field b 
  : self_A.a = self_A.a
  := eq_refl.
EndFamily.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.4\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module a_4 (self_A: EmptySig).
Definition a : nat := 1. End a_4.
Module Type a_5 (self_A: EmptySig).
Parameter (a : nat). End a_5.
Module Type a_6 := a_5 EmptyMod.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.3\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module b_7 (self_A: a_6).
Definition b 
  : self_A.a = self_A.a 
  := eq_refl. End b_7.

@@

% Concrete details of compilation of a family term data structure
Taking \cref{fig:plugin-example1} as an example: when the user interactively input the line "Field a : nat := ..", our plugin will translate this statement into the "Module a_9", "Module Type a_10, a_11" three components. {The "EmptySig" is just empty module type.} Then, these modules and module types will be generated and type-checked by Coq immediately. The computed "Module a_10" is then part of the context of "Field b" (as the module type of "self__A") for "b" to refer to. The internal representation of family "A" is simply the list of these ``compiled'' modules and module types. We can simply achieve incremental checking and overriding---for example, we want to override "Field a", we simply replace that compiled module with the new module inside the list. This compiled module type serve as an abstraction to let overriding possible.

@@

% Insert one pseudo-code example for explanation of the mechanism
% Use the natural number example
\begin{figure}
  \begin{minipage}[t]{0.3\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Family B.
  FInductive b : Set 
    := tt : b | ff : b.
  Family neg_handler. 
    Field tt : self_B.b 
      := self_B.ff.
    Field ff : self_B.b 
      := self_B.tt.
  EndFamily.
FRecursor neg 
  about self_B.b 
  motive (fun _ => self_B.b)
  using self_B.neg_handler
  by _rec.
EndFamily.

>>>>

\textbf{Family compiled into module.} This is the main idea of compiling family components into Coq's primitives: we compile family into coq module, family type into module type and context of a given term into parameters of the module. Unlike module, we need to faithfully reflect the \textit{late binding} nature of family : for each defining field $ .. \cL\sigma \vdash t : T $, we need to compile $t$ using ``universal-quantifier-wrapped term'' : $.. \vdash \lambda t : \forall \cL \sigma. T$. However, instead of using Coq's universal quantifier, we choose module and module parameters to achieve this wrapping---each family type in the context of the judgement will be one module parameter with the compiled module type, and thus each field of the family will compile into a functor parametrized by its context. Doing so we can also get rapid feedback from type checking when the users are defining each field interactively with Coq.
% Insert one pseudo-code example for explanation of the mechanism
\begin{figure}[!htb]
  \begin{minipage}[t]{0.25\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Family A.
Field a : nat 
  := 1.
Field b 
  : self_A.a = self_A.a
  := eq_refl.
EndFamily.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.4\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module a_4 (self_A: EmptySig).
Definition a : nat := 1. End a_4.
Module Type a_5 (self_A: EmptySig).
Parameter (a : nat). End a_5.
Module Type a_6 := a_5 EmptyMod.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.3\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module b_7 (self_A: a_6).
Definition b 
  : self_A.a = self_A.a 
  := eq_refl. End b_7.

@@

% Concrete details of compilation of a family term data structure
Taking \cref{fig:plugin-example1} as an example: when the user interactively input the line "Field a : nat := ..", our plugin will translate this statement into the "Module a_9", "Module Type a_10, a_11" three components. {The "EmptySig" is just empty module type.} Then, these modules and module types will be generated and type-checked by Coq immediately. The computed "Module a_10" is then part of the context of "Field b" (as the module type of "self__A") for "b" to refer to. The internal representation of family "A" is simply the list of these ``compiled'' modules and module types. We can simply achieve incremental checking and overriding as in \ref{langdesign-req0}---for example, we want to override "Field a", we simply replace that compiled module with the new module inside the list. This compiled module type serve as an abstraction to let overriding possible.

@@

% Insert one pseudo-code example for explanation of the mechanism
% Use the natural number example
\begin{figure}[!htb]
  \begin{minipage}[t]{0.3\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Family B.
  FInductive b : Set 
    := tt : b | ff : b.
  Family neg_handler. 
    Field tt : self_B.b 
      := self_B.ff.
    Field ff : self_B.b 
      := self_B.tt.
  EndFamily.
FRecursor neg 
  about self_B.b 
  motive (fun _ => self_B.b)
  using self_B.neg_handler
  by _rec.
EndFamily.

### 2022-06-20 lang-design.tex
Comments Reply.

\YZ{looks more like a 'design space' rather than 'language design' section}
When adapting Family Polymorphism into dependent types, we choose to
focus only on the essence of family (and inheritance)
structure in \citet{zm2017}, and thus a lot of unrelated features, like
Interface, will be removed. In this case, it will look like module with
late-binding and inheritance as code and proof reuse for module. We also
need it to have good compatibility with inductive types, because we
don't want to lose too much reasoning power of Coq (and its tactic
programming).

@@

% We want to reason about fields in a family.
\textlabel{Req (1)}{langdesign-req1}:\textbf{ we want to be able to
reason about fields in a family}, just like in a module, where a field
can reason about its former fields. More generally, we want to allow a
later field to be type-dependent on the former fields. In the example of
\cref{fig:STLC-example}, "subst", "has_type", "subst_lemma" all require
this feature.\YZ{rephrase Req to Challenges}

@@

% we want to be able to reason pattern matching as well
However, there is still difference between data recursion and induction
reasoning, because the former one is \textit{proof-relevant}. This
difference leads to another issue: \textlabel{Req (4)}{langdesign-req4}:
\textbf{we need to reason about (the computation about) extensible
pattern matching}, just like "subst_lemma" in \cref{fig:STLC-example}.
When proving "subst_lemma", we have to know "subst" is invariant on
"tt", i.e., \mintinline{Coq}{(subst i x tt) = tt}. This kind of
information requires exposing \textit{computational rules} from the
recursors.\YZ{What goes wrong if subst is not invariant or you do not
know whether subst is invariant?}

% we want tactic programming and certain level of automation
Finally, \textlabel{Req (5)}{langdesign-req5}: \textbf{we desire tactic
programming} to relieve us from the hassle of direct manipulation of
proof terms.\YZ{What about modular type checking?}

>>>>

\YZ{looks more like a 'design space' rather than 'language design' section}\EDJreply{Good point. Do I just change the section name?s}
When adapting Family Polymorphism into dependent types, we choose to
focus only on the essence of family (and inheritance)
structure in \citet{zm2017}, and thus a lot of unrelated features, like
Interface, will be removed. In this case, it will look like module with
late-binding and inheritance as code and proof reuse for module. We also
need it to have good compatibility with inductive types, because we
don't want to lose too much reasoning power of Coq (and its tactic
programming).

@@

First and foremost, \textlabel{Req (0)}{langdesign-req0}\textbf{we desire incremental(modular) type checking}. Notice that, the modularity here in Coq is a bit different from other language, because Coq is \textit{interactive} theorem proving so we actually need \textit{statement-wise} incremental type checking, not only for saving the compilation time, but also for the immediate feedback and type-checking for the Coq developer.

% We want to reason about fields in a family.
\textlabel{Req (1)}{langdesign-req1}:\textbf{ we want to be able to
reason about fields in a family}, just like in a module, where a field
can reason about its former fields. More generally, we want to allow a
later field to be type-dependent on the former fields. In the example of
\cref{fig:STLC-example}, "subst", "has_type", "subst_lemma" all require
this feature.\YZ{rephrase Req to Challenges}\EDJreply{But it is not really challenge because I didn't show their difficulty. It is more like requirements.}

@@

% we want to be able to reason pattern matching as well
However, there is still difference between data recursion and induction
reasoning, because the former one is \textit{proof-relevant}. This
difference leads to another issue: \textlabel{Req (4)}{langdesign-req4}:
\textbf{we need to reason about (the computation about) extensible
pattern matching}, just like "subst_lemma" in \cref{fig:STLC-example}.
When proving "subst_lemma", we have to know "subst" is invariant on
"tt", i.e., \mintinline{Coq}{(subst i x tt) = tt}. This kind of
information requires exposing \textit{computational rules} from the
recursors.\YZ{What goes wrong if subst is not invariant or you do not
know whether subst is invariant?}\EDJreply{Then it is unprovable because substitution lemma need to know this. You can just instantiate the whole lemma -- we want  
has_type (Γ; x ↦ A) tt T ->
has_type Γ u A ->
has_type Γ (subst i x tt) T. But this final goal requires to know the computation of subst.}

% we want tactic programming and certain level of automation
Finally, \textlabel{Req (5)}{langdesign-req5}: \textbf{we desire tactic
programming} to relieve us from the hassle of direct manipulation of
proof terms.\YZ{What about modular type checking?}\EDJreply{Good point! Let me add that.}
